---
title: "section1_DiD"
author: "Eline van Groningen, Paola Priante, Valery Maasdamme, Yuhu Wang"
date: "9/25/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, 
                      comment = "", fig.height=2, fig.width=4, fig.align = "center")
```

# Difference-in-Difference Analysis: Female Labor Participation 

```{r}
#Downloading the libraries, setting the working directory and importing the data set 
library(tidyverse)
library(stargazer)
library(dagitty)
library(gridExtra)
library(tinytex)
library(ggplot2)
library(tidyr)
library(sandwich)
library(multiwayvcov)

dir <- "/Users/valeriemaasdamme/Documents/BAM_ASP_A2" 
dirProg <- paste0(dir, "/programs/") 
dirData <- paste0(dir, "/Data/")

dfDiD <- read.csv(file=paste0(dirData, "DiD_dataset.csv"))
```

## Preparing and analyzing the dataset 
There is need to transform the dataset as it is already in the long-format, with observations in different rows for different time periods. There is no need to transform character to factor as all variables seem to be either numeric or integer. 

We have data from 1991 to 1996. In 1993, the EITC was implemented. A dummy variable *dPeriod* is created to indicate the time before and after the EITC policy. The years 1991 and 1992 take a value of 0, and 1993 and the years after 1993 take a value of 1.

The EITC policy only applies to women with children. Therefore, we need to compare two groups of women. A dummy variable *dChildren* is created to indicate the different groups: women with and without any children. Women with no children take a value of 0 (non-treated group), whereas women with 1 or more children take a value of 1 (treated group).

Furthermore, the EITC only applies to low-income workers. Therefore, women in both the treated and the non-treated group should be employed (*work* = 1). In order to estimate the DiD effect of the policy on annual earnings (*earn*) and annual family income (*finc*), we create a subset of women that are employed. 


```{r}
#no need to transform the dataset, already in the long format 
str(dfDiD) #all variable are numeric or integer, no need to tranform 


dfDiD$dPeriod = ifelse(dfDiD$year >= 1993, 1, 0) 
dfDiD$cChildren = ifelse(dfDiD$children >= 1, 1, 0)

dfDiD.sub <- subset(dfDiD, work=="1")
```

## 1 Plotting the dependent variables 
In order to present visual evidence of the difference-in-difference effects, the aggregate function in R is used to compute the averages of the three dependent variables per year per group. 

```{r}
#Earn
#6 years for both groups, total of 12 averages (average by year and children (0/1)
earn.agg = aggregate(dfDiD.sub$earn, list(dfDiD.sub$year, dfDiD.sub$cChildren == 1), FUN = mean, na.rm = TRUE) 
names(earn.agg) = c("Year","Children","Earn") #rename variables
#new variable with group name
earn.agg$Group[1:6] = "Women without children"
earn.agg$Group[7:12] = "Women with children"

Earn.plot <- qplot(Year, Earn, data=earn.agg, geom=c("point","line"), colour = Group,
  xlab="Year", ylab="Annual earnings") + 
  geom_vline(xintercept = 1993) + 
  theme_bw() 
ggsave(file="Earn.pdf", width=7, height=4)

#Finc
finc.agg = aggregate(dfDiD.sub$finc, list(dfDiD.sub$year, dfDiD.sub$cChildren == 1), FUN = mean, na.rm = TRUE) 
names(finc.agg) = c("Year","Children","Finc")
finc.agg$Group[1:6] = "Women without children"
finc.agg$Group[7:12] = "Women with children"

Finc.plot <- qplot(Year, Finc, data=finc.agg, geom=c("point","line"), colour = Group,
  xlab="Year", ylab="Annual Family Income") + 
  geom_vline(xintercept = 1993) +
  theme_bw() 
ggsave(file="Finc.pdf", width=7, height=4)

#Work
work.agg = aggregate(dfDiD$work, list(dfDiD$year, dfDiD$cChildren == 1), FUN = mean, na.rm = TRUE)
names(work.agg) = c("Year","Children","Work")

work.agg$Group[1:6] = "Women without children"
work.agg$Group[7:12] = "Women with children"

Work.plot <- qplot(Year, Work, data=work.agg, geom=c("point","line"), colour = Group,
  xlab="Year", ylab="Work")+ 
  geom_vline(xintercept = 1993) +
  theme_bw() 
ggsave(file="Work.pdf", width=7, height=4)

```

As can be seen in figure 1.1 and 1.2, there is an increase in the annual earnings and annual family income of women with children after the EITC was implemented. The annual earnings of women without any children seems more or less constant, with a relatively small increase in 1995.   

Figure 1.3 shows an increase of employed women with children after the implementation of the EITC. However, there seems to a decrease of employed women without children after the EITC was implemented. 

Based on the graphs above, the EITC seems to have a positive effect on annual earnings (*earn*), annual family income (*finc*) and rate of employment (*work*). 

## 2 Summary statistics of the dataset 

```{r}
stargazer(dfDiD, type = "text")
stargazer(dfDiD[, c("children", "finc", "earn", "age", "work", "unearn")], type = "text")
```

The distriptive statistics table above show that there are 13,746 observations in the dataset. The average number of children women in the dataset have is 1.193. The average annual earnings and annual family income are and 10,432.48 and 15,255.32 US Dollars respectively. The women in the dataset are on average 35 years old. Approximately half of the women in the dataset is employed, as the average employment rate equals 0.513. 

## 3 Difference-in-Difference 
In order to difference-in-difference effects in a matrix, the averages of the dependent variables are computed per group per period. 


```{r}
avgEarn <- ddply (dfDiD.sub, .(dPeriod, cChildren), summarise, 
                 avgEarn = mean(earn, na.rm=TRUE))

avgFinc <- ddply (dfDiD.sub, .(dPeriod, cChildren), summarise, 
                 avgFinc = mean(finc, na.rm=TRUE))

avgWork <- ddply (dfDiD, .(dPeriod, cChildren), summarise, 
                 avgWork = mean(work, na.rm=TRUE))

#Remodel the avg table from long to wide, add  additional row for the difference in averages and rename the rows
avgtable.Earn <- dcast (avgEarn, dPeriod ~ cChildren, value.var = "avgEarn")
avgtable.Earn <- rbind(avgtable.Earn, avgtable.Earn[2,]-avgtable.Earn[1,]) 
rownames(avgtable.Earn) <- c("Before", "After", "Difference")
colnames(avgtable.Earn) <- c("dPeriod", "Women without children (0)", "Women with children (1)")
avgtable.Earn[3, "dPeriod"] <- NA

avgtable.Finc <- dcast (avgFinc, dPeriod ~ cChildren, value.var = "avgFinc")
avgtable.Finc <- rbind(avgtable.Finc, avgtable.Finc[2,]-avgtable.Finc[1,]) 
rownames(avgtable.Finc) <- c("Before", "After", "Difference")
colnames(avgtable.Finc) <- c("dPeriod", "Women without children (0)", "Women with children (1)")
avgtable.Finc[3, "dPeriod"] <- NA

avgtable.Work <- dcast (avgWork, dPeriod ~ cChildren, value.var = "avgWork")
avgtable.Work <- rbind(avgtable.Work, avgtable.Work[2,]-avgtable.Work[1,]) 
rownames(avgtable.Work) <- c("Before", "After", "Difference")
colnames(avgtable.Work) <- c("dPeriod", "Women without children (0)", "Women with children (1)")
avgtable.Work[3, "dPeriod"] <- NA

stargazer(avgtable.Earn, summary=FALSE, align = TRUE, type="text", title = "Average Annual Earnings")
stargazer(avgtable.Finc, summary=FALSE, align = TRUE, type="text", title = "Average Indicator Annual Family Income")
stargazer(avgtable.Work, summary=FALSE, align = TRUE, type="text", title = "Average Indicator Work Status")

```

As can be seen in the tables above, the average annual earnings of employed women without children decreased with 1,076.42 US Dollars after the implementation of the EITC. However, the average annual earnings of woman with children increased with 451,54 US Dollars after the implementation of the EITC. The difference-in-difference effect of the EITC on annual earnings of employed women is 1,076.42 $+$ 451.538 $=$ 1,527.96. 

Similarly, the average annual family income of employed women without children decreased with 1,348.46 US Dollars after the implementation of the EITC, whereas the average family income of women with children increased with 455.36 US Dollars after the EITC was implemented. The difference-in-difference effect of the EITC on annual family income of employed women is 1,348.46 $+$ 455.360 $=$ 1,803.82. 

The average rate of employment for women without any children decreased with 0.005 after the EITC was implemented. On the other hand, the average rate of employment for women with children increased with 0.026 after the implementation of the EITC. The difference-in-difference effect of the EITC on the number of women that are employed is 0.005 $+$ 0.026 $=$ 0.031. 

## 4 Regression analysis 
The following difference-in-difference models are estimated:

$$Earn = \beta_0 + \beta_1 cChildren + \beta_2 dPeriod + \beta_3 cChildren dPeriod + \varepsilon_i$$
$$Finc = \beta_0 + \beta_1 cChildren + \beta_2 dPeriod + \beta_3 cChildren dPeriod + \varepsilon_i$$
$$Work = \beta_0 + \beta_1 cChildren + \beta_2 dPeriod + \beta_3 cChildren dPeriod + \varepsilon_i$$

```{r}

mdlEarn <- earn ~ cChildren + dPeriod + cChildren:dPeriod
rsltOLSEarn <- lm(mdlEarn, data=dfDiD.sub)

mdlFinc <- finc ~ cChildren + dPeriod + cChildren:dPeriod
rsltOLSFinc <- lm(mdlFinc, data=dfDiD.sub)

mdlWork <- work ~ cChildren + dPeriod + cChildren:dPeriod
rsltOLSWork <- lm(mdlWork, data=dfDiD)

stargazer(rsltOLSEarn, rsltOLSFinc, rsltOLSWork,
          intercept.bottom = FALSE, align = TRUE, no.space=TRUE, type="text")
```

Explain and interpret the findings. What is the effect of the policy introduction on the dependent variables? 
Based on the DiD regression in the table above, the difference-in-difference effect of the EITC on annual earnings is 1,527.959. This indicates an increase in annual earnings of employed women with atleast 1 child after the implementation of the EITC. However, this coefficient is not significant. 

Similarly, the effect of the EITC on annual family income based on the regression above, is 1,803.822. The coefficient is significant (p < 0.01). This indicates that the EITC increased the annual family income of employed woman with at least one child. 

The difference-in-difference effect of the EITC on the rate of employment of women is 0.031. The coefficient is significant (p < 0.01). This indicates that the EITC increased the rate of employed of women with at least one child. That, an increase of approximately 3%. 

## Control variables 
To control for other variables that may have an influence on the difference-in-difference effect of the EITC, control variables are added to the regressions of the three dependent variables. 

A possible explanation for the obtained regression results may be that state specific characteristics are not taken into account when comparing averages. Therefore, the state unemployment rate (*urate*) is added as a control variable. The rate of unemployment in states may have an effect on *work*, as women living in a state with a high unemployment rate are likely to be unemployed (*work* = 0) 

Another possible explanation for the results may be the unearned income of women (*unearn*). Women with a high income gained through other resources than work, e.g. through inheritance or investments, may be less inclined to work (*work* = 0). Therefore, *unearn* is added as a control variable. 
 
Lastly, the number of children of woman (*children*) is added as a control variable. It is expected that women with more children work less than women with a few children, as they may have a harder time balancing work and family life (*work* = 0). Consequently, this variable is added as a control variable. 

```{r} 
#Earn
mdl.control.earn <- earn ~ cChildren + dPeriod + cChildren:dPeriod + urate + unearn + children 
rsltOLS.control.earn <- lm(mdl.control.earn, data=dfDiD.sub)

#Finc
mdl.control.finc <- finc ~ cChildren + dPeriod + cChildren:dPeriod + urate + unearn + children
rsltOLS.control.finc <- lm(mdl.control.finc, data=dfDiD.sub)

#Work
mdl.control.work <- work ~ cChildren + dPeriod + cChildren:dPeriod + urate + unearn + children 
rsltOLS.control.work <- lm(mdl.control.work, data=dfDiD)

stargazer(rsltOLS.control.earn, rsltOLS.control.finc, rsltOLS.control.work, 
           intercept.bottom = FALSE, align = TRUE, no.space=TRUE, type="text") 

```

All control variables seem to be significant in the three regressions (p < 0.05). When controlling for the state unemployment rate, the unearned income and the number of children, the difference-in-difference effect of the EITC on *earn* decreased to 1,490.156. However, this coefficient is still not significant. 

Similarly, the difference-in-difference effect of the EITC on *finc* decreased from 1,803.822 to 1,490.156. However, after adding the control variables, the effect became insignificant. 

Furthermore, the difference-in-difference effect of the EITC on *work* increased from 0.031 to 0.035. The significance increased as well (p < 0.05). This indicates that if the state unemployment rate, the unearned income and the number of children are contolled for, the EITC led to a larger growth in employment of women with children (approximately 4%). 

## Robust standard errors 
The Breusch-Pagan test for heteroskedasticity was used to assess whether robust standard errors are necessary. The test was significant, which indicates a correlation between the predicted Y values and the error terms for all three DiD regressions. The detected heteroskedasticity is controlled for with White heteroskedasticity consistent standard errors and clustered standard errors (clustering based on *state*). The use of robust standard errors (White and clustered) does not have a critical effect on the statistical significance of the estimated DiD effect or the significance of the control variables. With the exception of the control variable *unearn*, that became less significant (p < 0.01) in the regression on *earn* after adding the White standard errors. Remarkably, the use of White standard errors leads to a smaller standard error compared to the base case for all three regressions, while the clustered standard errors remains the same as the base case. 

```{r}
#Test for heteroskedasticity 
rsltOLS.control.earn2 <- lm(mdl.control.earn, data=dfDiD.sub)
rsltOLS.control.finc2 <- lm(mdl.control.finc, data=dfDiD.sub)
rsltOLS.control.work2 <- lm(mdl.control.work, data=dfDiD)

# EARN
ggplot(data = data.frame(fit = fitted(rsltOLS.control.earn2), rsid = residuals(rsltOLS.control.earn2)),
       aes(fit, rsid)) +
  geom_point() + 
  stat_smooth(se = F) +
  theme_bw() +
  labs(x = "Results OLS Fitted") +
  labs(y = "Residuals")

lmtest::bptest(rsltOLS.control.earn2) 
# p < 0.001, indicates a significant correlation between the predicted values and the disturbances. 
# Heteroskedastiscity is detected.

#FINC
ggplot(data = data.frame(fit = fitted(rsltOLS.control.finc2), rsid = residuals(rsltOLS.control.finc2)),
       aes(fit, rsid)) +
  geom_point() + 
  stat_smooth(se = F) +
  theme_bw() +
  labs(x = "Results OLS Fitted") +
  labs(y = "Residuals")

lmtest::bptest(rsltOLS.control.finc2) 
# p < 0.001, indicates a significant correlation between the predicted values and the disturbances. 
# Heteroskedastiscity is detected.

#WORK
ggplot(data = data.frame(fit = fitted(rsltOLS.control.work2), rsid = residuals(rsltOLS.control.work2)),
       aes(fit, rsid)) +
  geom_point() + 
  stat_smooth(se = F) +
  theme_bw() +
  labs(x = "Results OLS Fitted") +
  labs(y = "Residuals")

lmtest::bptest(rsltOLS.control.work2) 
# p < 0.001, indicates a significant correlation between the predicted values and the disturbances 
# Heteroskedastiscity is detected

#Standard  errors 
seBasicEarn  <- sqrt(diag(vcov(rsltOLS.control.earn2)))
seWhiteEarn  <- sqrt(diag(vcovHC(rsltOLS.control.earn2, type="HC0")))
seClusterEarn <- sqrt(diag(vcov(rsltOLS.control.earn2, cluster="state"))) 
stargazer(rsltOLS.control.earn2, rsltOLS.control.earn2, rsltOLS.control.earn2, 
          se=list(seBasicEarn, seWhiteEarn, seClusterEarn), type="text") 
#No impact on the significance of the DiD effect, still insignificant 
#Standard error of seWhite seems smaller than basic and clustered 

seBasicFinc  <- sqrt(diag(vcov(rsltOLS.control.finc2)))
seWhiteFinc  <- sqrt(diag(vcovHC(rsltOLS.control.finc2, type="HC0")))
seClusterFinc <- sqrt(diag(vcov(rsltOLS.control.finc2, cluster="state"))) 
stargazer(rsltOLS.control.finc2, rsltOLS.control.finc2, rsltOLS.control.finc2, 
          se=list(seBasicFinc, seWhiteFinc, seClusterFinc), type="text")
#No impact on the significance of the DiD effect, still insignificant
#Standard error of seWhite seems smaller than basic and clustered 

seBasicWork  <- sqrt(diag(vcov(rsltOLS.control.work2)))
seWhiteWork  <- sqrt(diag(vcovHC(rsltOLS.control.work2, type="HC0")))
seClusterWork <- sqrt(diag(vcov(rsltOLS.control.work2, cluster="state"))) 
stargazer(rsltOLS.control.work2, rsltOLS.control.work2, rsltOLS.control.work2, 
          se=list(seBasicWork, seWhiteWork, seClusterWork), type="text")
#No impact on the significance of the DiD effect, all three significant (p<0.05).
#Standard error for all three remains the same

stargazer(rsltOLS.control.earn2, rsltOLS.control.earn2, rsltOLS.control.earn2, 
          rsltOLS.control.finc2, rsltOLS.control.finc2, rsltOLS.control.finc2,
          rsltOLS.control.work2, rsltOLS.control.work2, rsltOLS.control.work2, 
          se=list(seBasicEarn, seWhiteEarn, seClusterEarn,
                  seBasicFinc, seWhiteFinc, seClusterFinc,
                  seBasicWork, seWhiteWork, seClusterWork), type="text") 
```


